---
title: "Version comparison"
author: "S. Kitanovski, K. Wollek"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Version comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "#>",
                      fig.height = 5,
                      fig.width = 5,
                      fig.align = "center")
```

# Case study - Extended version comparison

To demonstrate the advantage of *version 3* of the `r Biocpkg("ClustIRR")` 
algorithm, we will conduct the same case study as integrated in the user manual
vignette.

We start by loading the needed packages. *knitr*, *ggplot2* and *ggrepel* are
used to display the results.

```{r case_study}
# load packages
library(ClustIRR)
library(knitr)
library(ggplot2)
library(ggrepel)
```

## Input data
A randomly sampled part of the reference dataset will be used as a sample 
dataset. To test the algorithm against ground truth, the sample dataset will be
artificially enriched with the motif *RQWW* and clonally expanded with the two
sequences *CATSRAAKPDGLAALETQYF* and *CATSRAAKPDGLAALSTQYF*.

The package provided `CD8` dataset is loaded as an example reference dataset. 
This set of 10.000 β-chains is derived from the T cell receptors of human naive
CD8+ T cells was published as part of the Gliph 2\(^2\) algorithm.


```{r input_data}
# load package input data
data("CD8")

# set the random seed for sampling reproducibility
set.seed(987)

# sample 500 sequences from the reference dataset as sample dataset
data_sample <- data.frame(CDR3b = CD8[sample(x = 1:nrow(CD8), 
                                             size = 500, 
                                             replace = FALSE),])

# artificially enrich motif 'RQWW' inside sample dataset
substr(x = data_sample$CDR3b[1:20], start = 6, stop = 9) <- "RQWW"

# add the artificial clonal expansion of two sequences to the sample dataset
clones <- data.frame(CDR3b = rep(x = c("CATSRAAKPDGLAALETQYF",
                                       "CATSRAAKPDGLAALSTQYF"),
                                 times = 15))
data_sample <- rbind(data_sample, clones)

```

## Clustering

Only the sample and reference
dataset must be provided as input to perform clustering with the default 
values. For demonstration purposes, we will only look for motifs of length 
`ks` = 4 and compute outputs for all three versions of the algorithm, i.e.
*version 3*, *version 2* and *version 1*. Each time we will trim three flank 
amino acids on each side of all sequences.

```{r clustering}
# ClustIRR version 3
clustirr_output_v3 <- cluster_irr(data_sample = data_sample,  
                                  data_ref = CD8,
                                  version = 3,
                                  ks = 4,
                                  cores = 1,
                                  control = list(
                                    B = 1000,
                                    global_max_dist = 1,
                                    local_max_fdr = 0.05,
                                    local_min_ove = 2,
                                    local_min_o = 3,
                                    trim_flank_aa = 3,
                                    low_mem = FALSE,
                                    global_pairs = NULL))
# ClustIRR version 2
clustirr_output_v2 <- cluster_irr(data_sample = data_sample,  
                                  data_ref = CD8,
                                  version = 2,
                                  ks = 4,
                                  cores = 1,
                                  control = list(
                                    B = 1000,
                                    global_max_dist = 1,
                                    local_max_fdr = 0.05,
                                    local_min_ove = 2,
                                    local_min_o = 3,
                                    trim_flank_aa = 3,
                                    low_mem = FALSE,
                                    global_pairs = NULL))
# ClustIRR version 1
clustirr_output_v1 <- cluster_irr(data_sample = data_sample,  
                                  data_ref = CD8,
                                  version = 1,
                                  ks = 4,
                                  cores = 1,
                                  control = list(
                                    B = 1000,
                                    global_max_dist = 1,
                                    local_max_fdr = 0.05,
                                    local_min_ove = 2,
                                    local_min_o = 3,
                                    trim_flank_aa = 3,
                                    low_mem = FALSE,
                                    global_pairs = NULL))


```
Warnings tell us that CDR3 sequences of the reference datasets shorter than 
6 amino acids were trimmed completely before local clustering was performed.
This happens, because we set the `trim_flank_aa` parameter to 3, so at each 
side of all CDR3 sequences 3 aa got cut off, effectively trimming sequences 
shorter than 6 aa completely. As long as our sample datasets are not cut, this 
should be fine.

## Output

### Local motif clustering output - version 3

Inspecting the local motif clustering of *version 3* shows that, in total, 20 
motifs were marked as enriched. 

6 motifs relating to the 20 times inserted motif get found. The inserted motif
*RQWW* gets found 20 times (`f_sample` column). Five other motifs are 
combinations of the first three (*RQW* <-> *QRQW*, *YRQW*) or last three 
(*QWW* <-> *QWWA*, *QWWG*, *QWWS*) amino acids of the inserted motif with the 
respective sequences it was inserted in.

14 motifs related to the clonally expanded sequences *CATSRAAKPDGLAALETQYF* 
and *CATSRAAKPDGLAALSTQYF* are found. Some of them are found in both sequences 
(for example *AAKP*, found 30 times), some of them only in one of the two (for
example *AALE*, found 15 times). Notice that several of the newly found motifs 
were also present in the reference dataset as parts of other sequences 
(`f_ref` > 0), and in this case, by chance, one of them (*GLAA*) was contained 
in the sample before the artificial enhancement, leading to a count of 31. 

If the random seed value changed, slightly different motif counts 
could appear through interaction with sampled motifs of the reference set. 
In any case, the inserted motifs and the motifs found in the clonally 
expanded sequences are guaranteed to be detected by the algorithm *version 3*.


```{r local_motifs_output_version_3}
# extract local motifs version 3
local_motifs_v3 <- clustirr_output_v3$clust$CDR3b$local$m

# display passed motifs
kable(local_motifs_v3[local_motifs_v3$pass == TRUE,])

```

### Local motif clustering output - version 2

Inspecting the local motif clustering of *version 2* shows that, in total, 6 
motifs were marked as enriched. The 20 times inserted motif *RQWW* gets found 
20 times(`f_sample` column). The other five motifs are all combinations of the 
first three (*RQW* <-> *QRQW*, *YRQW*) or last three (*QWW* <-> *QWWA*, *QWWG*,
*QWWS*) amino acids of the inserted motif with the respective sequences it was 
inserted in.

Because of the redundancy reduction performed in *version 2*, no motifs of the 
2 clonal expanded sequences *CATSRAAKPDGLAALETQYF* and *CATSRAAKPDGLAALSTQYF* 
are found, which both were inserted 15 times into the sample data.


```{r local_motifs_output_version_2}
# extract local motifs version 2
local_motifs_v2 <- clustirr_output_v2$clust$CDR3b$local$m

# display passed motifs
kable(local_motifs_v2[local_motifs_v2$pass == TRUE,])
```

### Local motif clustering output - version 1

Inspecting the local motif clustering of *version 1* shows that, in total, 12 
motifs were marked as enriched. The 20 times inserted motif *RQWW* gets found 
20 times (`f_sample` column). Five other motifs are combinations of the first 
three amino acids (*RQW* <-> *QRQW*, *YRQW*) or last three amino acids (*QWW* 
<-> *QWWA*, *QWWG*, *QWWS*) amino acids of the inserted motif with the 
respective sequences it was inserted in.

Because of the redundancy reduction performed in *version 1*, no motifs of the 
2 clonal expanded sequences *CATSRAAKPDGLAALETQYF* and *CATSRAAKPDGLAALSTQYF* 
are found, which both were inserted 15 times into the sample data.

*Version 1* uses bootstrapping to compute enrichment (*version 3* and 
*version 2* use Fisher's exact test, see man-files for details). 
This can induce noise, as repeated samples are drawn randomly, and also limits
the p-value precision.

The remaining 6 motifs *SYRQ*, *SNRA*, *SLVS*, *LVSG*, *GGSY* and *GFNT* are
found in this case, but if the random seed is changed, other motifs appear
as enriched.

```{r local_motifs_output_version_1}
# extract local motifs version 1
local_motifs_v1 <- clustirr_output_v1$clust$CDR3b$local$m

# display passed motifs
kable(local_motifs_v1[local_motifs_v1$pass == TRUE,])
```

### Global CDR3 sequence clustering output

Inspecting the global motif clustering shows the same output for all versions.
The first pair of CDR3b sequences (*CASSYVLRAADGYTF* and *CASSYVLRAADGDTF*) 
were present in the reference dataset and were randomly sampled into the 
example dataset, while the second pair (*CATSRAAKPDGLAALSTQYF* and 
*CATSRAAKPDGLAALETQYF*) represents the artificially enhanced clonal expansion.

Again, if the sampling were repeated with a different seed, other random 
pairs of the reference dataset could be sampled, but the two added sequences
are guaranteed to be found by the algorithm.


```{r global_motifs_output}
# global motifs version 1
kable(clustirr_output_v1$clust$CDR3b$global)

# global motifs version 2
kable(clustirr_output_v2$clust$CDR3b$global)

# global motifs version 3
kable(clustirr_output_v3$clust$CDR3b$global)
```
### Graph output

As mentioned in the user manual vignette, graph output is planned to be added
in future `r Biocpkg("ClustIRR")` versions.

# Session Info

```{r package_version}
packageVersion("ClustIRR")
```


```{r session_info}
sessionInfo()
```
