---
title: "Benchmarking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{benchmarking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
require(gliphR)
require(ggplot2)
require(patchwork)
```

This vignette benchmarks the gliphR package versions. <br>
Input data consists of an artificially expanded CD8 reference repertoire, 
taken from (turboGliph, source to be inserted). For noise testing, identical data is input as reference and sample. <br>
Input parameters for all gliphR versions are matched as closely
as possible for comparability. <br>
Clustering results are compared by OvE value. <br>
Results to be interpreted once finished (tba)

TODO: add more meaningful graphs once gliphR 1 is fixed


# Create input data

An artificial clonal expansion of 100 vs 1 sequences of a HIV specific
CDR3b sequence is created as input.

```{r create_input_data}

# set seed for reproducibility 
set.seed(1234)

# HIV specific CDR3b sequence, taken from last research project
# TODO insert correct source / database
# CASSETGGTYEQYF TRBV6-1 	TRBJ2-7 	
hiv_expansion <- c(CDR3b = "CASSETGGTYEQYF", 
                   TRBV = "TRBV6-1", 
                   TRBJ = "TRBJ2-7 ")

data("hs_CD8_ref")
data_ref <- hs_CD8_ref[, 1:3]

# insert single CDR3b sequence into reference repertoire
data_ref <- rbind(data_ref, hiv_expansion)

# sample 5000 sequences from reference repertoire
data_sample <- hs_CD8_ref[sample(x = 1:nrow(data_ref), 
                                 size = 5000, replace = FALSE), 1:3]


# create artificial expansion of 100 sequences
for(i in 0:99) {
  data_sample <- rbind(data_sample, hiv_expansion)
}



```


# Prepare input parameters

```{r create_parameters}

cores <- parallel::detectCores()

# gliphR v1 + v2 + v3
ks <- c(2, 3, 4)
B <- 1000
control_input <- list(
  B = B,
  global_max_dist = 1,
  local_max_fdr = 0.05,
  local_min_ove = 2,
  local_min_o = 3,
  trim_flank_aa = 3,
  low_mem = FALSE,
  global_pairs = NULL)


```

# Cluster with gliphR versions

```{r run_gliph}
out_v1 <- gliph(data_sample = data_sample,
                data_ref = data_ref,
                ks = ks,
                cores = cores,
                version = 1,
                control = control_input)

out_v2 <- gliph(data_sample = data_sample,
                data_ref = data_ref,
                ks = ks,
                cores = cores,
                version = 2,
                control = control_input)


out_v3 <- gliph(data_sample = data_sample,
                data_ref = data_ref,
                ks = ks,
                cores = cores,
                version = 3,
                control = control_input)

# put in the same data for ref and sample
out_v1_i <- gliph(data_sample = data_sample,
                  data_ref = data_sample,
                  ks = ks,
                  cores = cores,
                  version = 1,
                  control = control_input)

out_v2_i <- gliph(data_sample = data_sample,
                  data_ref = data_sample,
                  ks = ks,
                  cores = cores,
                  version = 2,
                  control = control_input)


out_v3_i <- gliph(data_sample = data_sample,
                  data_ref = data_sample,
                  ks = ks,
                  cores = cores,
                  version = 3,
                  control = control_input)

```


# Compare results

```{r compare_results}

# prepare for visualization
e_v1 <- out_v1$clust$CDR3b$local$m
colnames(e_v1)[which(colnames(e_v1) == 'ove')]     <- 'OvE_gliphR1'
e_v1_i <- out_v1_i$clust$CDR3b$local$m
colnames(e_v1_i)[which(colnames(e_v1_i) == 'ove')] <- 'OvE_gliphR1_identical'
e_v2 <- out_v2$clust$CDR3b$local$m
colnames(e_v2)[which(colnames(e_v2) == 'ove')]     <- 'OvE_gliphR2'
e_v2_i <- out_v2_i$clust$CDR3b$local$m
colnames(e_v2_i)[which(colnames(e_v2_i) == 'ove')] <- 'OvE_gliphR2_identical'
e_v3 <- out_v3$clust$CDR3b$local$m
colnames(e_v3)[which(colnames(e_v3) == 'ove')]     <- 'OvE_gliphR3'
e_v3_i <- out_v3_i$clust$CDR3b$local$m
colnames(e_v3_i)[which(colnames(e_v3_i) == 'ove')] <- 'OvE_gliphR3_identical'


```


## Identical vs different input & version comparison

```{r identical_vs_different, fig.width=10, fig.height=10, fig.align='center'}

# gliphR v1 test versus identical input
c_v1_i <- merge(e_v1, e_v1_i, by = "motif", all = TRUE)
c_v1_i[is.na(c_v1_i)] <- -1
g_v1 <- ggplot(c_v1_i, aes(x = OvE_gliphR1_identical, y = OvE_gliphR1)) +
  geom_point() +
  geom_abline(linewidth = 0.5, color = "gray") +
  ggtitle("gliphR v1 - identical versus different input")

# gliphR v2 test versus identical input
c_v2_i <- merge(e_v2, e_v2_i, by = "motif", all = TRUE)
c_v2_i[is.na(c_v2_i)] <- -1
g_v2 <- ggplot(c_v2_i, aes(x = OvE_gliphR2_identical, y = OvE_gliphR2)) +
  geom_point() +
  geom_abline(linewidth = 0.5, color = "gray") +
  ggtitle("gliphR v2 - identical versus different input")

# gliphR v3 test versus identical input
c_v3_i <- merge(e_v3, e_v3_i, by = "motif", all = TRUE)
c_v3_i[is.na(c_v3_i)] <- -1
g_v3 <- ggplot(c_v3_i, aes(x = OvE_gliphR3_identical, y = OvE_gliphR3)) +
  geom_point() +
  geom_abline(linewidth = 0.5, color = "gray") +
  ggtitle("gliphR v3 - identical versus different input")

# gliphR v1 vs gliphR v2
c_v1_2 <- merge(e_v1, e_v2, by = "motif", all = TRUE)
c_v1_2[is.na(c_v1_2)] <- -1
g_gl1_gl2 <- ggplot(c_v1_2, aes(x = OvE_gliphR1, y = OvE_gliphR2)) +
  geom_point() +
  geom_smooth(method = lm , color = "blue", fill = "red") +
  geom_abline(linewidth = 0.5, color = "gray") +
  ggtitle("gliphR v1 vs gliphR v2")

# gliphR v1 vs gliphR v3
c_v1_3 <- merge(e_v1, e_v3, by = "motif", all = TRUE)
c_v1_3[is.na(c_v1_3)] <- -1
g_gl1_gl3 <- ggplot(c_v1_3, aes(x = OvE_gliphR1, y = OvE_gliphR3)) +
  geom_point() +
  geom_smooth(method = lm , color = "blue", fill = "red") +
  geom_abline(linewidth = 0.5, color = "gray") +
  ggtitle("gliphR v1 vs gliphR v3")

# gliphR v2 vs gliphR v3
c_v2_3 <- merge(e_v2, e_v3, by = "motif", all = TRUE)
c_v2_3[is.na(c_v2_3)] <- -1
g_gl2_gl3 <- ggplot(c_v2_3, aes(x = OvE_gliphR2, y = OvE_gliphR3)) +
  geom_point() +
  geom_smooth(method = lm , color = "blue", fill = "red") +
  geom_abline(linewidth = 0.5, color = "gray") +
  ggtitle("gliphR v2 vs gliphR v3")

( g_v1 + g_v2 + g_v3 ) / (g_gl1_gl2 + g_gl1_gl3 + g_gl2_gl3)


```
